Think again?\\ Optimal mental simulation tracks problem difficulty

### 2015-01-26 manuscript.tex
More work on modeling and fitting

%% What we're going to do and how we're going to do it
The plan of the paper is as follows. First, we formalize our hypothesis for how people respond to the question, ``will the ball go through the hole?'', specifying both a simulation-based model as well as the SPRT strategy. Second, we describe two behavioral experiments, one in which we asked people to respond to the question of ``will the ball go through the hole?'', and a related experiment that we used to determine the parameters for the simulation-based model. Third, we describe the results of the experiments, and compare people's judgments and response times to those from the model. Finally, we discuss the implications of our results on the broader, underlying question: how do people use mental simulation?

Consider the task in Figure \ref{fig:experiment}, in which people observe a blue ball moving around in a box, and are then asked, ``will the ball go through the hole?''. How do people make this type of judgment? We hypothesize that they predict where the ball will go by running noisy physical simulations, and that they run a variable number of simulations on each trial according to the \textit{sequential probability ratio test}, or SPRT \cite{wald1947sequential}. In this section, we lay out the formalism for both the simulation-based model and the SPRT strategy.

Specifically, \citeA{Smith2012} asked people to catch a ball like the one in Figure \ref{fig:experiment} using a paddle that could move up and down along the $y$-axis. In order to predict where people would place the paddle on each trial, their model incorporated uncertainty in its physical simulations through four parameters: perceptual uncertainty over the position of the ball ($\sigma_p$), uncertainty in the initial direction of velocity ($\kappa_v$), ongoing uncertainty in the direction of movement ($\kappa_m$), and uncertainty in the bounce angle ($\kappa_b$).

In the best possible case, the agent takes infinite samples and chooses the maximum \emph{a posteriori} (MAP) hypothesis with probability $p$. In practice, the agent cannot take infinite samples. Thus, to determine when to stop sampling (i.e., what the value of $N$ is), we use the \emph{sequential probability ratio test} \cite{wald1947sequential}. When using a SPRT strategy, the agent accumulates $Y_N=\sum_{i=1}^N 2X_i-1$, and stops either when $Y_N=T$ or $Y_N=-T$, for some threshold number of samples $T$. Independent of the number of samples actually taken, the probability of choosing $H_1$ given that $H_1$ is the MAP hypothesis is:
\begin{equation}
\Pr[\hat{H}(Y_N)=H_1\,|\,H_1]=\frac{p^T}{p^T+(1-p)^T},
\label{eq:pr-choose-h1}
\end{equation}

Given the threshold, the probability of taking $N$ samples and then choosing $H_1$ is:
\begin{multline}
\Pr[N,H_1\,|\,T,p]=\left(\frac{2^N}{2T}\right)p^{\frac{N+T}{2}}(1-p)^{\frac{N-T}{2}}\cdot{}\\
\sum_{\nu=1}^{2T-1}\cos\left(\frac{\nu\pi}{2T}\right)^{N-1}\sin\left(\frac{\nu\pi}{2T}\right)\sin\left(\frac{\nu\pi}{2}\right)
\end{multline}
for $N\geq T$ and where $N$ is of the same parity as $T$ \cite[ch.~XIV, eq. 5.7]{Feller:1968ut}. Because we are dealing with binary hypotheses, the marginal probability of $N$ samples is then:
\begin{equation}
\Pr[N\,|\,T,p]=\Pr[N,H_1\,|\,T,p]+\Pr[N,H_1\,|\,T,1-p].
\label{eq:pr-n}
\end{equation}
and so the expected number of samples for a particular decision is:
\begin{equation}
\mathbb{E}[N\,|\,T,p]=\sum_{N=T}^\infty N\cdot{}\Pr[N\,|\,T,p]
\end{equation}

To combine the ``Noisy Newton'' hypothesis with the SPRT strategy, we needed to compute the probability, $p$, that the ball goes in the hole. To do this, we simulated $M=10000$ samples from the model. From these samples, we estimated a truncated normal distribution along the wall with the hole in it, and then computed the probability density lying within the region of the hole.

\TODO{talk about standard deviation adjustment}

\TODO{talk about adjusting for the number of bounces -- or should this be in the results?}

\section{Experiment 1}

@@

As shown in Figure \ref{fig:pct-vs-rt} (right subplot), we found a clear speed-accuracy tradeoff in participants responses and response times. In particular, participants were on average faster to respond on trials that they were more certain about, and slower to respond to trials that they were unsure about. This tradeoff mirrors the same tradeoff found in SPRT: decisions for which $p\approx0.5$ take much longer to make, because it takes longer to get to one threshold versus the other.

@@

\TODO{summarize results showing that experimen 2 replicates Smith \& Vul (2014), and talk about fitting the model parameters}

@@

We fit participant responses and response times to the SPRT model with different values of $T$, and found that $T=2$ to have the best fit. Participant responses were well correlated with the SPRT model ($r=0.81$, 95\% CI [0.77, 0.84]), and slightly less so with a SPRT model with $T=2$ ($r=0.78$, 95\% CI [0.74, 0.81]). There was also a significant correlation with resonse time for the $T=2$ SPRT model ($r=0.35$, 95\% CI [0.27, 0.43]). However, it turns out that the number of bounces is a much better predictor of people's response times than the model itself.

>>>>

%% Plan of the paper
The plan of the paper is as follows. First, we describe an experiment in which we asked participants to respond to the question of, ``will the ball go through the hole?'', and analyze people's accuracy and response times. Second, we formalize our hypothesis of how people make judgments in this experiment by specifying a combination of a simulation-based model with the SPRT strategy. Third, we describe a second experiment based on the experiment from \citeA{Smith2012} that was used to fit the parameters of our model. Then, we compare the model's predictions to people's responses in Experiment 1. Finally, we discuss the implications of our results on the broader, underlying question: how do people use mental simulation?

\section{Experiment 1: Will the ball go in the hole?}

\begin{figure*}[t]
    \begin{center}
        \includegraphics[width=\textwidth]{figures/human-pct-vs-rt.pdf}
        \caption{\textbf{Response times as a function of response.} Left subplot: each bar shows the proportion of participants saying that the ball will go in the hole for a particular feedback type ($x$-axis) and hole size (color). Middle subplot: like the left subplot, but the $y$-axis shows bootstrapped logarithmic means of response times. Right subplot: each point corresponds to a different stimulus, feedback type, and hole size.  The $x$-axis is the proportion of participants saying the ball will go in the hole, and the $y$ axis is the logarithmic mean response time. \TODO{fix legends} \TODO{right subplot shouldn't be speed vs. accuracy}}
        \label{fig:pct-vs-rt}
    \end{center}
\end{figure*}

In order to determine whether people take multiples samples---and if they vary the number of samples---we designed an experiment in which people had to determine whether a ball would go into a hole (see Figure \ref{fig:experiment}). In particular, we designed the trials in this experiment such that some were harder than others.

@@

As shown in Figure \ref{fig:pct-vs-rt} (right subplot), we found a clear effect of certainty on people's response times. In particular, participants were on average faster to respond on trials that they were more certain about, and slower to respond to trials that they were unsure about. This tradeoff mirrors the same tradeoff found in SPRT: decisions for which $p\approx0.5$ are slower, because it takes longer to get to one threshold versus the other. Thus, it is clear that people are not spending a fixed amount of time thinking about each trial; rather, they spend more time thinking about the ``hard'' trials than the ``easy'' trials.

How do people make the types of judgments that they did in Experiment 1? We hypothesized that they predict where the ball will go by running noisy physical simulations, and that they run a variable number of simulations on each trial according to the \textit{sequential probability ratio test}, or SPRT \cite{wald1947sequential}. In this section, we lay out the formalism for both the simulation-based model and the SPRT strategy.

Specifically, \citeA{Smith2012} asked people to catch a ball like the one in Figure \ref{fig:experiment} using a paddle that could move up and down along the $y$-axis. In order to predict where people would place the paddle on each trial, their model incorporated uncertainty in its physical simulations through five parameters: perceptual uncertainty over the position of the ball ($\sigma_p$), uncertainty in the initial direction of velocity ($\kappa_v$), ongoing uncertainty in the direction of movement ($\kappa_m$), uncertainty in the bounce angle ($\kappa_b$), and a ``center bias'' reflecting a bias for people to respond more towards the center of the $y$-axis ($\sigma_0$).

In the best possible case, the agent takes infinite samples and chooses the maximum \emph{a posteriori} (MAP) hypothesis with probability $p$. In practice, the agent cannot take infinite samples. Thus, to determine when to stop sampling (i.e., what the value of $N$ is), we use the \emph{sequential probability ratio test} \cite{wald1947sequential}. When using a SPRT strategy, the agent accumulates $Y_N=\sum_{i=1}^N 2X_i-1$, and stops either when $Y_N=T$ or $Y_N=-T$, for some threshold number of samples $T$.

\subsubsection{Modeling responses} Independent of the number of samples actually taken, the probability of choosing $H_1$ given that $H_1$ is the MAP hypothesis is:
\begin{equation}
\Pr[\hat{H}(Y_N)=H_1\,|\,H_1]=\frac{p^T}{p^T+(1-p)^T},
\label{eq:pr-choose-h1}
\end{equation}

We can compute the probability, $p$, that the ball goes in the hole by running simulations from the model.

\subsubsection{Modeling response times} Given the threshold, the probability of taking $N$ samples and then choosing $H_1$ is:
\begin{multline}
\Pr[N,H_1\,|\,T,p]=\left(\frac{2^N}{2T}\right)p^{\frac{N+T}{2}}(1-p)^{\frac{N-T}{2}}\cdot{}\\
\sum_{\nu=1}^{2T-1}\cos\left(\frac{\nu\pi}{2T}\right)^{N-1}\sin\left(\frac{\nu\pi}{2T}\right)\sin\left(\frac{\nu\pi}{2}\right)
\end{multline}
for $N\geq T$ and where $N$ is of the same parity as $T$ \cite[ch.~XIV, eq. 5.7]{Feller:1968ut}. Because we are dealing with binary hypotheses, the marginal probability of $N$ samples is then:
\begin{equation}
\Pr[N\,|\,T,p]=\Pr[N,H_1\,|\,T,p]+\Pr[N,H_1\,|\,T,1-p].
\label{eq:pr-n}
\end{equation}
and so the expected number of samples for a particular decision is:
\begin{equation}
\mathbb{E}[N\,|\,T,p]=\sum_{N=T}^\infty N\cdot{}\Pr[N\,|\,T,p]
\end{equation}

Assuming every sample takes the same amount of time, response times as predicted by the model should be directly proportional to $\mathbb{E}[N\,|\,T,p]$. However, we found the number of bounces to be an extremely strong predictor of response time in Experiment 1: each additional bounce adds a constant amount of time to the response. To account for this, our final response time equation incorporated the number of bounces, $B$, as follows:
\begin{equation}
t = \beta_0 + (\beta_1 + \beta_2\cdot{}\mathbb{E}[B]) \cdot{}\mathbb{E}[N\,|\,T,p]
\label{eq:rt}
\end{equation}
where $\beta_0$, $\beta_1$, and $\beta_2$ are fitted coefficients. \TODO{figure}

@@

\TODO{summarize results showing that experimen 2 replicates Smith \& Vul (2014)}

We fit the model in two steps. First, we fit $\sigma_p$, $\kappa_v$, $\kappa_m$, $\kappa_b$, and $\sigma_0$ to particant's responses in Experiment 2. \TODO{explain the procedure by which we did this fitting} Using this procedure, we found the best fitting parameters to be $\sigma_p=31.02$, $\kappa_v=255.60$, $\kappa_m=502850.41$, $\kappa_b=50.42$, and $\sigma_0=167.06$.

From the fitted model, we can estimate the probability that the ball goes in the hole ($p$), with one adjustment. Specifically, it is not clear how many samples people take to solve the paddle task. If they are taking a fixed number of samples, $M$,\footnote{We assume that in the paddle experiment, people are taking a fixed number of samples, as they only have a small, fixed amount of time to respond.} then we can estimate the true standard deviation of their simulations as $\sigma_{sims} = \sigma_{judgments} / \sqrt{M}$. As $M$ is unknown, we treat it as an additional free parameter in our model, adjusting the standard deviation of the model's responses by a fixed constant, $\sigma_{adj}=\sqrt{M}$.

We fit $\sigma_{adj}$ and $T$ using a grid search to maximize the correlation between SPRT response times (Equation \ref{eq:rt}) and actual response times in Experiment 1. To compute Equation \ref{eq:rt}, we ran $n=10000$ samples from the model, and from these samples (as well as the $\sigma_{adj}$ adjustment) estimated a truncated normal distribution along the wall with the hole in it. From this truncated normal, we computed the value of $p$ as the the probability density lying within the region of the hole. From this procedure, we found the best fitting parameters to be $\sigma_{adj}=0.7$ \TODO{what is the exact value?} and $T=2$.

\TODO{what about fitting the coefficients from Equation \ref{eq:rt}?}

@@

Participant responses were well correlated with the SPRT model ($r=0.81$, 95\% CI [0.77, 0.84]), and slightly less so with a SPRT model with $T=2$ ($r=0.78$, 95\% CI [0.74, 0.81]). There was also a significant correlation with resonse time for the $T=2$ SPRT model ($r=0.35$, 95\% CI [0.27, 0.43]). However, it turns out that the number of bounces is a much better predictor of people's response times than the model itself.
